---
title: "GIS-03 - Mapping with ggplot2 and tamp"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction
This notebook expands mapping to two new packages - `ggplot2` and `tmap`. We'll focus on thematic choropleth mapping today.

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# tidyverse packages
library(dplyr)        # data wrangling
library(ggplot2)      # plotting

# spatial packages
library(mapview)      # preview spatial data
library(sf)           # spatial data tools
library(tigris)       # access TIGER/Line shapefile data
library(tmap)         # mapping

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
```

## Load Data
First, we need to load up our data. This is review from the first session, and so we've pre-filled the code for you. We'll automatically re-project the data as well (also review):

```{r load-data}
# polygon data 1 - St. Louis Neighborhood Population
nhood <- st_read(here("data", "STL_DEMOS_Nhoods", "STL_DEMOS_Nhoods.shp")) %>%
  st_transform(crs = 4326)

# polygon data 2 - St. Louis COVID Cases by ZIP Code
covid <- st_read(here("data", "daily_snapshot_regional.geojson"))
```

## Download a Bit of Extra Data
We'll use the same process as last week to create boundary data for both the city and the region:

```{r load-counties}
region <- counties(state = 29) %>%
  select(GEOID, NAMELSAD) %>%
  filter(GEOID %in% c("29099", "29183", "29189", "29510")) %>%
  st_transform(crs = 4326)

city <- filter(region, GEOID == "29510")
```

## Simple Maps with `ggplot2`
### Basic Mapping of Geometric Objects
`ggplot2` is the premier graphics package for `R`. It is an incredibly powerful visualization tool that increasingly supports spatial work and mapping. The basic `ggplot2` workflow requires chaining together functions with the `+` sign. 

We'll start by creating a `ggplot2` object with the `ggplot()` function, and then adding a "geom", which provides `ggplot2` instructions on how our data should be visualized. We can read these like paragraphs:

1. First, we create an empty `ggplot2` object, **then**
2. we add the `nhood` data and visualize its geometry.

```{r ggplot2-nhoodSimple}
ggplot() +
  geom_sf(data = nhood, fill = "#bababa")
```

You can see empty spaces where there are major parks - if we wanted to give these a background color, we could add the `city` layer under our `nhood` layer. We can also add the `city` layer again on top to give the city border a pronounced outline. `ggplot2` relies on layering different geoms to produce complicated plots. We can assign each geom a specific set of aesthetic characteristics and use data from different objects.

```{r ggplot2-nhoodSimple2}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = NA) +
  geom_sf(data = nhood, fill = "#bababa") +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75)
```

Now it is your turn - re-create this process but map zip codes using the `covid` data and use the regional county boundaries in `region`:

```{r ggplot2-covidSimple}

```

### Mapping Quantities with `ggplot2`
If we wanted to start to map data instead of just the geometric properties, we would specify an "aesthetic mapping" using `mapping= aes()` in the geom of interest. Here, we create a fill that is the product of taking the population in 2017 and normalizing it by square kilometers as we did in the `leaflet` section above. We provide additional instructions about how our data should be colored with the `scale_fill_distiller()` function, which gives us access to the `RColorBrewer` palettes.

```{r ggplot2-nhoodFull}
ggplot() +
  geom_sf(data = city, fill = "#ffffff", color = NA) +
  geom_sf(data = nhood, mapping = aes(fill = pop17/(AREA/1000000))) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "Greens", trans = "reverse", name = "Population per\nSquare Kilometer") +
  labs(
    title = "Population Density (2017)",
    subtitle = "Neighborhoods in St. Louis, MO",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() 
```

Replicate this process, using the `case_rate` column in `covid` to plot the already normalized numbers of cases per 1,000 people in each zip code:

```{r ggplot2-covidFull}

```

## Map Layouts with `tmap`
`tmap` uses a similar logic to `ggplot2` - it layers elements on top of each other to produce maps. It is dedicated to working with spatial data, however, and has some features that `ggplot2` does not. 

### Basic Mapping of Geometric Objects
We'll start with a basic map that, like we have previously, just display the geometry of the city's neighborhoods. Similar to `ggplot2`, functions are chained together with the `+` sign. We can read these like paragraphs:

1. First, we take the `nhood` data, **then**
2. we create our `tmap` layer out of its shape, **then**
3. we add a fill using our layer, **then**
4. we add borders using our layer.
  
```{r tmap-simple}
nhood %>%
  tm_shape() +
    tm_fill() +
    tm_borders() 
```

### Mapping Quantities with `tmap`
Like `ggplot2`, we can plot quantities using the `tm_polygons()` function. The `palette` argument accepts names of both `RColorBrewer` and `viridis` palettes. `tmap` also contains a number of tools for creating map layouts that "feel" more like desktop GIS outputs.

```{r tmap-nhoodFull}
tm_shape(city) +
  tm_fill(fill = "#ebebeb") + 
  tm_shape(nhood) +
  tm_polygons(col = "pop17", 
              palette = "viridis", 
              style = "jenks",
              convert2density = TRUE,
              title = "Population per\nSquare Kilometer") +
  tm_shape(city) +
  tm_borders(lwd = 2) +
  tm_scale_bar() +
  tm_layout(
    title = "Population Density (2017)",
    frame = FALSE,
    legend.outside = TRUE,
    legend.position = c("left", "bottom"))
```

Now, it is your turn to repeat this process. Create a map layout for the `covid` data, mapping `case_rate` again. (Hint: you won't need to set `convert2density` to `TRUE`)

```{r tmap-covidFull}

```

## Saving Maps
There are two processes for saving maps, one for `ggplot2` and one for `tmap`. `ggplot2` uses `ggsave()` while `tmap` uses `tmap_save()`.
